---
title: "GDPandEducation_CaseStudy1"
author: "Georges Michel, Jack Nelson, Nicole Shaver, Nathan Tuttle"
date: "February 25, 2017"
output: html_document
Keep_md: TRUE

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#set the working directory
# knitr::opts_knit$set(root.dir = "C:\\Users\\nmsha\\Documents\\SMU\\DoingDataScience\\homework\\CaseStudy1")
knitr::opts_knit$set(root.dir = "..\\")
#ensure all required packages are installed, and if not, install them
list.of.packages <- c("downloader")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
library(downloader)
```

## Introduction

In this study we will examine the Gross Domestic Product (GDP) rankings for countries across the world and analyze the relationship between GDP and their education levels. We will order the countries by GDP and evaluate the average GDP ranking for different education groups, visualize the GDP data for all countries by income group, and examine the relationship between income group and GDP. The analysis will utilize GDP and education data from the data-catalog at the world bank web site.


## Code used to download, tidy and merge the data
```{r download_tidy_merge}
dir<-getwd()
source(paste(dir,"/source/download.R",sep=""))
source(paste(dir,"/source/cleanup_ED_GDP.R",sep=""))
source(paste(dir,"/source/merge_ED_GDP.R",sep=""))
```

## Analysis Results

1. After merging the GDP and education datasets based on country code, 224 of the IDs match between the two datasets, as shown in the dataframe below, which has 224 rows after the matching between the education data and the GDP data is complete and NA's are excluded:
```{r count_rows}
##use the str command to describe the number of observations.
##rows where the country code did not match were exluded by this line in merge_ED_GDP.R
combo2<-subset(combo,!is.na(Long.Name) & !is.na(GDP.Country))
str(combo2)
```

2. The country with the 13th lowest GDP (determined by sorting the data in ascending order by GDP, placing the United States last) is St. Kitts and Nevis. There were 21 NA values in the GDP column when sorting, but they were all at the end.
```{r sorting}
#sort in order of GDP
sorted_GDP<-combo2[order(combo2$GDP),]
#pick out the 13th row from the dataframe, just the columns needed to answer the question
GDP_13<-sorted_GDP[13,c(1,2,3,4)]
##display the 13th row
GDP_13
#count the NA in the GDP column
sum(is.na(sorted_GDP$GDP))
```

3. The average GDP ranking for the "High Income: OECD" group is 32.96667. The average GDP ranking for the "High Income: nonOECD" group is 91.91304. There were 14 NAs in the GDP ranking column. 
```{r question 3}
# Create subset dataframe containing only 'High income: OECD' and 
# 'High income: nonOECD' income group countries
combo2.HighIncome <- subset(combo2, combo2$Income.Group == 'High income: OECD' | combo2$Income.Group == 'High income: nonOECD')

# Take the average GDP.ranking by income group of the subset with NA values removed
aggregate(combo2.HighIncome$GDP.ranking~combo2.HighIncome$Income.Group, combo2.HighIncome, mean, na.rm = TRUE)
#count the number of NA values in GDP ranking
sum(is.na(combo2.HighIncome$GDP.ranking))
```

4. Plotting the GDP for all countries, coloring by Income Group, we note that xxxxxxxxx.
```{r cars5, echo=TRUE}
getwd()
```
5. Breaking the GDP ranking into 5 separate quantile groups, and then summarizing the Income Groups, we see, as shown in the table below, that 2 countries are in the "Lower Middle Income" Group but among the 38 nations with the highest GDP.
```{r question 5}

# Group GDP by ranking for values that are not N/A
GDPRanking <- subset(combo2,!is.na(GDP.ranking))

combo2$GDP.quant.group[combo2$GDP.ranking <20] <-"G0_to_20"
combo2$GDP.quant.group[combo2$GDP.ranking >=20 & combo2$GDP.ranking <40 ] <-"G20_to_40"
combo2$GDP.quant.group[combo2$GDP.ranking >=40 & combo2$GDP.ranking <60 ] <-"G40_to_60"
combo2$GDP.quant.group[combo2$GDP.ranking >=60 & combo2$GDP.ranking <80 ] <-"G60_to_80"
combo2$GDP.quant.group[combo2$GDP.ranking >=80 & combo2$GDP.ranking <100 ] <-"G80_to_100"

data.frame(table(combo2$Income.Group,combo2$GDP.quant.group))


```

## Conclusion
In this study of GDP and Education levels, using data from the world bank, we find that there are 224 countries from which we can make conclusions about the GDP and education levels. The country with the 13th lowest GDP is St. Kitts and Nevis.